---
title: "Data aggregating notebook for Biketown PDX"
author: "Raymond Ford (raymond.anthony.ford@gmail.com)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_notebook:
    toc: yes
  html_document:
    df_print: paged
    toc: yes
---

This notebook will take the interim data and transform it into a form that separates `Casual` and `Subscriber` users.

# Preliminaries
Before we begin this notebook we will need the following function and `R` package.

## Needed function
This function is based on a solution provided from [Stack Overflow](https://stackoverflow.com/questions/5200437/aggregating-hourly-data-into-daily-aggregates), and we have adapted it to accompish what we need it to do.

```{r}
totals <- function(x){
  # This function will take a numerical vector x and output the sum of x and
  # the mean of x.
  # ----------
  # INPUT
  # x := a numerical vector.
  # ----------
  # OUTPUT
  # A data object with columns containing the sum and mean of x.
  c(sum=sum(x), mean=mean(x))
}
```

## Needed packages
In order to aggregate the data we will take advantage of the syntactic sugar provided by the `zoo` package avialable in `R`.

```{r}
require(zoo)
```

## Bringing in the data
We begin by bringing the data into `R`, and output a summary of the data using `summary()`.

```{r}
int.dat <- read.csv("https://raw.githubusercontent.com/raford/pdxbikes/master/data/interim/interim.csv",
                    header=TRUE, sep=",")
summary(int.dat)
```

We next subset the data into two distinct groups: one for `Casual` users and one for the `Subscriber` group---both based on the payment plan used.
```{r}
cas.dat <- subset(int.dat, PaymentPlan == "Casual")[, -1] # Payment plan in first column
sub.dat <- subset(int.dat, PaymentPlan == "Subscriber")[, -1]
rm(int.dat)
```

# Aggregate the data
We next construct two separate data sets: one containing the data for the `Casual` users and one containing the `Subscriber` users, and then write each to their own CSV file. The code used to create the datasets for each of these classes is below, and labeled appropriately.

## Casual users
```{r}
cas.dat$StartDate <- as.Date(cas.dat$StartDate)
Date <- seq.Date(from=min(cas.dat$StartDate), to=max(cas.dat$StartDate), "days")
cas.dat <- cas.dat[, -c(2, 3, 4, 7)]
cas.dat.agg <- read.zoo(cas.dat, header=TRUE, aggregate=totals)
cas.dat.agg <- cbind(Date, as.data.frame(cas.dat.agg))
write.csv(cas.dat.agg, file="~/GoogleDrive/pdxbikes/pdxbikes/data/aggregate/casual.csv",
          row.names=FALSE)
rm(cas.dat, cas.dat.agg)
```

## Subscribers
```{r}
sub.dat$StartDate <- as.Date(sub.dat$StartDate)
Date <- seq.Date(from=min(sub.dat$StartDate), to=max(sub.dat$StartDate), "days")
sub.dat <- sub.dat[, -c(2, 3, 4, 7)]
sub.dat.agg <- read.zoo(sub.dat, header=TRUE, aggregate=totals)
sub.dat.agg <- cbind(Date, as.data.frame(sub.dat.agg))
write.csv(sub.dat.agg, file="~/GoogleDrive/pdxbikes/pdxbikes/data/aggregate/subscriber.csv",
          row.names=FALSE)
rm(sub.dat, sub.dat.agg)
```

# Future work

1. Perform some exploratory data analysis (EDA) for these data sets.
2. Create `R` scripts that will create these data sets.

# Session information
Below you will find the output from `sessionInfo()` to assist in reproducing the work shown in this notebook.

```{r}
sessionInfo()
```

